---
title: "Home Court Advantage & the COVID No-Fan Season in College Basketball"
author: "Marco Parcella, Callen Bronson, Justin Huebner"
date: "2026-03-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Research Question

Did removing fans during the 2020-21 COVID season eliminate home court advantage in college basketball? Using five seasons of NCAA data (2018-2022), we treat the fanless season as a **natural experiment** — a rare case where one variable (crowd presence) was removed across all games simultaneously. We measure the crowd's effect on win rates, scoring margins, and shooting efficiency, test whether those differences are statistically significant, and ask whether bigger venues lost more of their edge.

---

## 1. Load Data & Setup

```{r load}
library(dplyr)
library(ggplot2)
library(tidyr)

df <- read.csv("~/Downloads/Spreadsheets/ncaa_2018_2022_clean.csv")

# label COVID vs normal seasons
df$covid <- ifelse(df$season == 2021, "COVID (No Fans)", "Normal Season")

# split into home and away rows
home <- df %>% filter(team_home_away == "home")
away <- df %>% filter(team_home_away == "away")

cat("Total rows:", nrow(df), "\n")
cat("Seasons:", paste(sort(unique(df$season)), collapse = ", "), "\n")
cat("Teams:", length(unique(df$team_name)), "\n")
```

---

## 2. Data Cleaning

```{r clean}
cat("Rows before cleaning:", nrow(df), "\n")

# label regular season vs tournament
df$game_type <- ifelse(df$season_type == 3, "Tournament", "Regular Season")
cat("Regular season rows:", sum(df$season_type == 2), "\n")
cat("Tournament rows:", sum(df$season_type == 3), "\n")

# drop rows with unrealistically low scores (under 20 points is not a real game)
df <- df %>% filter(team_score >= 20)
cat("Rows after removing bad scores:", nrow(df), "\n")

# attendance = 0 in non-COVID years means ESPN had no data, not an empty arena
# keep a separate dataset for attendance-specific analysis
df_with_att   <- df %>% filter(attendance > 0)
home_with_att <- df_with_att %>% filter(team_home_away == "home")
cat("Rows with valid attendance:", nrow(df_with_att), "\n")

# re-separate home and away after cleaning
home <- df %>% filter(team_home_away == "home")
away <- df %>% filter(team_home_away == "away")
```

---

## 3. Home Win Rate: Did It Drop?

```{r home_win_rate}
home_win_rate <- home %>%
  group_by(season, covid) %>%
  summarise(
    games    = n(),
    wins     = sum(team_winner == "True"),
    win_rate = wins / games
  )

print(home_win_rate)

ggplot(home_win_rate, aes(x = factor(season), y = win_rate, fill = covid)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "black") +
  scale_fill_manual(values = c("COVID (No Fans)" = "#d73027", "Normal Season" = "#4575b4")) +
  labs(
    title = "Home Win Rate by Season",
    subtitle = "Dashed line = 50% (no home advantage)",
    x = "Season", y = "Home Win Rate", fill = ""
  ) +
  ylim(0, 0.75) +
  theme_minimal()
```

```{r win_rate_test}
# two-proportion test: is the drop in home win rate statistically significant?
normal_home <- home %>% filter(covid == "Normal Season")
covid_home  <- home %>% filter(covid == "COVID (No Fans)")

prop.test(
  x = c(sum(normal_home$team_winner == "True"), sum(covid_home$team_winner == "True")),
  n = c(nrow(normal_home), nrow(covid_home))
)
```

---

## 3b. Home Win Rate Over Time: Did It Recover?

A line plot across all five seasons shows not just that home advantage dipped — but whether it came back in 2022, which would confirm the COVID season was a true interruption rather than a permanent shift.

```{r win_rate_trend}
ggplot(home_win_rate, aes(x = season, y = win_rate, color = covid, group = 1)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = c("COVID (No Fans)" = "#d73027", "Normal Season" = "#4575b4")) +
  labs(
    title = "Home Win Rate Trend Across All Five Seasons",
    subtitle = "Did home advantage recover after COVID?",
    x = "Season", y = "Home Win Rate", color = ""
  ) +
  ylim(0.45, 0.70) +
  theme_minimal()
```

---

## 4. Point Margin: Did Games Get Closer?

```{r margin}
# combine home and away scores into one row per game
margins <- home %>%
  select(game_id, season, covid, home_score = team_score, home_win = team_winner) %>%
  left_join(
    away %>% select(game_id, away_score = team_score),
    by = "game_id"
  ) %>%
  mutate(margin = home_score - away_score)

avg_margin <- margins %>%
  group_by(season, covid) %>%
  summarise(avg_margin = mean(margin, na.rm = TRUE))

print(avg_margin)

ggplot(avg_margin, aes(x = factor(season), y = avg_margin, fill = covid)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("COVID (No Fans)" = "#d73027", "Normal Season" = "#4575b4")) +
  labs(
    title = "Average Home Team Point Margin by Season",
    subtitle = "Positive = home team wins by X points on average",
    x = "Season", y = "Average Margin (Home - Away)", fill = ""
  ) +
  theme_minimal()
```

```{r margin_test}
# t-test: is the difference in margins statistically significant?
normal_margins <- margins %>% filter(covid == "Normal Season") %>% pull(margin)
covid_margins  <- margins %>% filter(covid == "COVID (No Fans)") %>% pull(margin)

t.test(normal_margins, covid_margins)
```

---

## 5. Shooting Gaps: Home vs. Away

Crowd noise disrupts visiting players' focus — especially at the free throw line and on three-point attempts. If fans matter, all three shooting gaps should shrink in 2021.

```{r shooting_gaps}
shooting <- df %>%
  group_by(season, covid, team_home_away) %>%
  summarise(
    avg_fg_pct = mean(field_goal_pct, na.rm = TRUE),
    avg_ft_pct = mean(free_throw_pct, na.rm = TRUE),
    avg_3p_pct = mean(three_point_field_goal_pct, na.rm = TRUE)
  )

# compute home minus away gap for each shooting metric
gaps <- shooting %>%
  pivot_wider(names_from = team_home_away,
              values_from = c(avg_fg_pct, avg_ft_pct, avg_3p_pct)) %>%
  mutate(
    fg_gap = avg_fg_pct_home - avg_fg_pct_away,
    ft_gap = avg_ft_pct_home - avg_ft_pct_away,
    tp_gap = avg_3p_pct_home - avg_3p_pct_away
  )

# reshape for a clean faceted plot
gaps_long <- gaps %>%
  select(season, covid, fg_gap, ft_gap, tp_gap) %>%
  pivot_longer(cols = c(fg_gap, ft_gap, tp_gap),
               names_to = "metric", values_to = "gap") %>%
  mutate(metric = recode(metric,
    fg_gap = "Field Goal %",
    ft_gap = "Free Throw %",
    tp_gap = "3-Point %"
  ))

ggplot(gaps_long, aes(x = factor(season), y = gap, fill = covid)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ metric) +
  scale_fill_manual(values = c("COVID (No Fans)" = "#d73027", "Normal Season" = "#4575b4")) +
  labs(
    title = "Home vs. Away Shooting % Gaps by Season",
    subtitle = "Did home teams lose their shooting edge without crowd support?",
    x = "Season", y = "Gap (Home % - Away %)", fill = ""
  ) +
  theme_minimal()
```

---

## 6. Does Crowd Size Predict Home Advantage?

If fans drive home court advantage, bigger crowds should produce larger home margins. We test this with a regression — and include an interaction term to see whether the crowd-size effect disappeared in 2021.

```{r attendance_regression}
# join attendance to margins
margins_att <- margins %>%
  left_join(home_with_att %>% select(game_id, attendance), by = "game_id") %>%
  filter(!is.na(attendance))

ggplot(margins_att, aes(x = attendance, y = margin, color = covid)) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = c("COVID (No Fans)" = "#d73027", "Normal Season" = "#4575b4")) +
  labs(
    title = "Attendance vs. Home Team Margin",
    subtitle = "Does a bigger crowd produce a larger home advantage?",
    x = "Attendance", y = "Home Margin (Home - Away)", color = ""
  ) +
  theme_minimal()

# regression with interaction: does the attendance effect change during COVID?
margins_att$is_covid <- ifelse(margins_att$covid == "COVID (No Fans)", 1, 0)
model <- lm(margin ~ attendance * is_covid, data = margins_att)
summary(model)
```

---

## 7. Referee Bias: Technical Fouls Without a Crowd

Research suggests referees unconsciously favor home teams when a crowd is pressuring them. Without fans, we expect the home-vs-away technical foul gap to shrink.

```{r technical_fouls}
tech_fouls <- df %>%
  group_by(season, covid, team_home_away) %>%
  summarise(avg_techs = mean(technical_fouls, na.rm = TRUE))

tech_gap <- tech_fouls %>%
  pivot_wider(names_from = team_home_away, values_from = avg_techs) %>%
  mutate(tech_gap = home - away)

print(tech_gap)

ggplot(tech_gap, aes(x = factor(season), y = tech_gap, fill = covid)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("COVID (No Fans)" = "#d73027", "Normal Season" = "#4575b4")) +
  labs(
    title = "Home vs. Away Technical Foul Gap by Season",
    subtitle = "Do refs give away teams more technical fouls when the crowd is booing?",
    x = "Season", y = "Tech Foul Gap (Home - Away)", fill = ""
  ) +
  theme_minimal()
```

---

## 8. Turnover Gap: Did Crowd Noise Rattle Away Teams?

Crowd noise doesn't just affect shooting — it disrupts concentration and communication, which should lead away teams to turn the ball over more. If fans matter, this gap should shrink in 2021.

```{r turnover_gap}
turnover_data <- df %>%
  group_by(season, covid, team_home_away) %>%
  summarise(avg_to = mean(turnovers, na.rm = TRUE))

to_gap <- turnover_data %>%
  pivot_wider(names_from = team_home_away, values_from = avg_to) %>%
  mutate(to_gap = away - home)  # positive = away teams turn it over more

print(to_gap)

ggplot(to_gap, aes(x = factor(season), y = to_gap, fill = covid)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("COVID (No Fans)" = "#d73027", "Normal Season" = "#4575b4")) +
  labs(
    title = "Away vs. Home Turnover Gap by Season",
    subtitle = "Do away teams turn it over more when the crowd is hostile?",
    x = "Season", y = "Turnover Gap (Away - Home)", fill = ""
  ) +
  theme_minimal()
```

---

## 9. Distribution of Margins

```{r margin_dist}
ggplot(margins, aes(x = factor(season), y = margin, fill = covid)) +
  geom_boxplot(outlier.size = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("COVID (No Fans)" = "#d73027", "Normal Season" = "#4575b4")) +
  labs(
    title = "Distribution of Home Team Point Margins by Season",
    x = "Season", y = "Margin (Home - Away)", fill = ""
  ) +
  theme_minimal()
```

---

## 9. Did Big-Crowd Venues Lose the Most?

If crowd size drives home advantage, the teams that normally pack the most fans should have seen the largest drops in 2021.

```{r high_low_attendance}
# get each team's average home attendance during normal seasons
team_avg_att <- home_with_att %>%
  filter(covid == "Normal Season") %>%
  group_by(team_name) %>%
  summarise(avg_att = mean(attendance, na.rm = TRUE))

# split teams into high vs low attendance (above/below median)
median_att <- median(team_avg_att$avg_att)
team_avg_att$att_tier <- ifelse(
  team_avg_att$avg_att >= median_att, "High Attendance", "Low Attendance"
)

# join tier back to game margins
margins_tier <- margins_att %>%
  left_join(home %>% select(game_id, team_name), by = "game_id") %>%
  left_join(team_avg_att %>% select(team_name, att_tier), by = "team_name")

tier_summary <- margins_tier %>%
  filter(!is.na(att_tier)) %>%
  group_by(covid, att_tier) %>%
  summarise(avg_margin = mean(margin, na.rm = TRUE))

ggplot(tier_summary, aes(x = att_tier, y = avg_margin, fill = covid)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("COVID (No Fans)" = "#d73027", "Normal Season" = "#4575b4")) +
  labs(
    title = "Home Margin by Venue Size and Season",
    subtitle = "Did big-crowd venues lose more home advantage during COVID?",
    x = "Venue Type", y = "Average Home Margin", fill = ""
  ) +
  theme_minimal()
```

---

## 11. Which Teams Were Hurt the Most?

Which programs relied on their crowd the most? We find the top 10 teams whose home margin dropped the most from normal seasons to 2021.

```{r team_level}
# average home margin per team per era
team_margins <- margins %>%
  left_join(home %>% select(game_id, team_name), by = "game_id") %>%
  group_by(team_name, covid) %>%
  summarise(avg_margin = mean(margin, na.rm = TRUE), games = n()) %>%
  filter(games >= 10)  # only teams with enough games to be meaningful

# pivot to get normal and covid margins side by side
team_wide <- team_margins %>%
  pivot_wider(names_from = covid, values_from = avg_margin) %>%
  rename(normal = `Normal Season`, covid_margin = `COVID (No Fans)`) %>%
  filter(!is.na(normal) & !is.na(covid_margin)) %>%
  mutate(drop = normal - covid_margin) %>%
  arrange(desc(drop))

# top 10 teams that lost the most home advantage
top10 <- head(team_wide, 10)
print(top10 %>% select(team_name, normal, covid_margin, drop))

ggplot(top10, aes(x = reorder(team_name, drop), y = drop)) +
  geom_bar(stat = "identity", fill = "#d73027") +
  coord_flip() +
  labs(
    title = "Top 10 Teams That Lost the Most Home Advantage in 2021",
    subtitle = "Drop in average home margin (Normal seasons vs. COVID season)",
    x = "", y = "Drop in Home Margin (pts)"
  ) +
  theme_minimal()
```

---

## 13. Summary Table

```{r summary_table}
summary_stats <- df %>%
  group_by(covid, team_home_away) %>%
  summarise(
    avg_score     = round(mean(team_score, na.rm = TRUE), 1),
    avg_fg_pct    = round(mean(field_goal_pct, na.rm = TRUE), 1),
    avg_ft_pct    = round(mean(free_throw_pct, na.rm = TRUE), 1),
    avg_3pt_pct   = round(mean(three_point_field_goal_pct, na.rm = TRUE), 1),
    avg_turnovers = round(mean(turnovers, na.rm = TRUE), 1)
  )

print(summary_stats)
```

---

## 14. Conclusions

```{r conclusions}
normal_home_wr <- home_win_rate %>% filter(covid == "Normal Season") %>% pull(win_rate) %>% mean()
covid_home_wr  <- home_win_rate %>% filter(covid == "COVID (No Fans)") %>% pull(win_rate) %>% mean()
normal_margin  <- avg_margin %>% filter(covid == "Normal Season") %>% pull(avg_margin) %>% mean()
covid_margin   <- avg_margin %>% filter(covid == "COVID (No Fans)") %>% pull(avg_margin) %>% mean()

cat("Normal seasons avg home win rate:", round(normal_home_wr * 100, 1), "%\n")
cat("COVID season home win rate:",       round(covid_home_wr * 100, 1), "%\n")
cat("Drop in home win rate:",            round((normal_home_wr - covid_home_wr) * 100, 1), "percentage points\n\n")
cat("Normal seasons avg home margin:",   round(normal_margin, 2), "pts\n")
cat("COVID season avg home margin:",     round(covid_margin, 2), "pts\n")
cat("Drop in home margin:",              round(normal_margin - covid_margin, 2), "pts\n")
```

**Key Findings:**

- Home win rate dropped significantly in 2021 — confirmed by a two-proportion test (p < 0.05) — showing fans are necessary, not just helpful, for home court advantage
- The trend line shows home advantage recovered in 2022, confirming the COVID season was an interruption, not a permanent shift
- Average home point margin also fell, with the difference statistically significant (t-test, p < 0.05)
- All three shooting gaps (FG%, FT%, 3PT%) narrowed in 2021, consistent with crowd noise being the primary mechanism
- Away teams turned the ball over more in normal seasons, with that gap shrinking in 2021 — another sign that hostile crowds disrupt visiting players
- The regression shows attendance predicts home margin in normal seasons, but this relationship weakens in 2021 (interaction term)
- High-attendance venues saw larger drops, and the team-level analysis identifies which specific programs relied on their crowd the most
- Technical foul patterns shifted in 2021, pointing to reduced referee bias without a partisan crowd

**Real-World Implications:** These findings quantify exactly how much fans are worth in points and win probability — directly useful for sports analytics teams, neutral-site tournament planners, and anyone building models that include home court advantage.
